Lifschitz Realizability for Homotopy Type Theory

### 2019-11-02 draft.tex
semi-final draft, intro still missing

\usepackage{tikz-cd}
%\usepackage{answers}
\usepackage{biblatex}
\addbibresource{bibliography.bib}
\usepackage{setspace}
%\usepackage{graphicx}
\usepackage{enumitem}
%\usepackage{tikz-cd}
%\usepackage{todonotes}
%\usepackage{multicol}
%\usepackage{mathrsfs}
%\usepackage[margin=1in]{geometry} 
\usepackage{hyperref}% http://ctan.org/pkg/hyperref
\usepackage{ stmaryrd }
\usepackage{amsmath}
\usepackage{cleveref}% http://ctan.org/pkg/cleveref
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{siunitx}
\usepackage{ textcomp }
%\usepackage{ wasysym }
\usepackage{bookmark}
%\usepackage{accents}
\usepackage[parfill]{parskip}

\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\nin}{\not \in}

\DeclareMathOperator{\sech}{sech}
\DeclareMathOperator{\csch}{csch}

\newenvironment{theorem}[2][Theorem]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{definition}[2][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{proposition}[2][Proposition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{lemma}[2][Lemma]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{exercise}[2][Exercise]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{solution}[2][Solution]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{problem}[2][Problem]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{question}[2][Question]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
\newenvironment{corollary}[2][Corollary]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}

\usepackage{xcolor}
\newcommand\todo[1]{\textcolor{red}{#1}}

@@

\newtheorem{thm}{Theorem}[section]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{axiom}[thm]{Axiom}
\newtheorem{defn}[thm]{Definition}
\theoremstyle{definition}
\newtheorem{rem}[thm]{Remark}
\newtheorem{ill}[thm]{Illustration}
\newtheorem{exmp}[thm]{Example}

\begin{document}
 %\pagenumbering{gobble}
% --------------------------------------------------------------
%                         Start here
% --------------------------------------------------------------

\title{Draft}%replace with the appropriate homework number
\author{Dimitrios Koutsoulis\\ %replace with your name
11838639} %if necessary, replace with your course title

\begin{titlepage}
\hypersetup{pageanchor=false}
\maketitle
\thispagestyle{empty}
\end{titlepage}
\hypersetup{pageanchor=true}
%Below is an example of the problem environment
\chapter{Type Theory}
\section{Introduction}
Martin-L\"of's intensional \textit{Type Theory} is a formal language and deductive system, that is self sufficient in the sense that it need not be formulated as a collection of axioms on top of some other formal system like First Order Logic, instead its deductive system can be built on top of its own formal language.

Central to Type Theory is the notion of \textit{Type}. Every term $a$ in Type Theory we come across, must lie in some type $A$, which we denote as $a : A$. 
To avoid impredicativity we assume a countable hierarchy of universes $\mathcal{U}_0,\mathcal{U}_1,\mathcal{U}_2,\ldots$ which is cummulative i.e. every universe includes all previous universes and their types. 
While working in type theory we usually simplify our view of this hierarchy and pick some $\mathcal{U}$, of arbitrary index that we do not specify, to use as our workspace. 
%Note that the relation $:$ is transitive, so $a : A$ and $A : B$ imply $a : B$. 
%This induces the cummulative hierarchy shape of the universe of types $\mathcal{U}$, wherein every level of the universe includes all lower levels and their types.

@@

\section{Type Construction Operations}\label{informalTT}
We have the following list of types at our disposal.
\begin{itemize}
	\item Given types $A, B : \mathcal{U}$ we can define the type $A \rightarrow B : \mathcal{U}$ of \textbf{functions} from $A$ to $B$. 
		We can use $\lambda$-abstraction to construct elements of this type. $\lambda x. \Phi$ lies in $A \rightarrow B$ iff for $a : A$ we have $\Phi[a/x] : B$. 
		For $f : A \rightarrow B$ and $a : A$ we have that the application of $f$ on $a$, denoted as $f(a)$ or $f\;a$, lies in $B$, so $f\;a : B$.

		Functions whose type is of the form $A \rightarrow \mathcal{U}$ i.e. the codomain is $\mathcal{U}$ are called \textit{families of types}. 
		They are of special interest because they can be viewed as types themselves; types indexed by terms of the domain type. 
		So if $B : A \rightarrow \mathcal{U}$ is such a family of types, then its inhabitants would be the collection of inhabitants of all $B(a)$'s for all $a  : A$. 
		This last sentence is not sanctioned by formal type theory and only serves to make the introduction of the first dependent type, the \textit{dependent function type}, easier.
	\item Given some type $A : \mathcal{U}$ and a family of types $B$ over $A$, $B : A \rightarrow \mathcal{U}$, we have the type of \textbf{dependent functions} or \textbf{dependent products} (which should not be viewed as the dependent version of the product type defined later on) $$\prod_{a : A} B(a)$$
	where for $f : \prod_{a : A} B(a)$ and $x : A$ we have $f(x) : B(x)$. 
	As in the case of non-dependent functions, we can use lambda abstraction to construct elements of a dependently-typed function type. 
	That way, $\lambda x.\Phi$ lies in $\prod_{a:A}B(a)$ iff $B$ is of the form $\lambda y.\Psi$ and for all $a:A$ we have $\Phi[a/x] : \Psi[a/y]$. 
	\item Given $A, B : \mathcal{U}$ we can define the \textbf{product} type $A \times B : \mathcal{U}$. 
	For $a : A$ and $b : B$ we have the pair $(a,b) : A \times B$. 
	We also have the projection functions 
	$$\mathtt{pr}_1 : A \times B \rightarrow A : (a,b) \mapsto a$$
	$$\mathtt{pr}_2 : A \times B \rightarrow B : (a,b) \mapsto b$$
	We sometimes use the notation $x.\mathtt{pr}_i$ to refer to $\mathtt{pr}_i\; x$. 
	We also make use of the alias $\mathtt{fst}$ for $\mathtt{pr}_1$ and $\mathtt{snd}$ for $\mathtt{pr}_2$.

@@

	The induction principle is the following
	$$\mathtt{ind}_{\sum_{a:A}B(a)} : \prod_{C : \sum_{a:A}B(a) \rightarrow \mathcal{U}}\Big( \prod_{a : A}\prod_{b : B(a)}C((a,b)) \Big)\prod_{x : \sum_{a:A}B(a)}C(x)$$
	$$\mathtt{ind}_{\sum_{a:A}B(a)}(C, f,(a,b) ) :\equiv f(a)(b)$$
	\item Given $A, B : \mathcal{U}$ we can construct the \textbf{coproduct} type $A + B$. 
	We can construct elements of $A + B$ using the functions
	$$\mathtt{inl} : A \rightarrow A+B$$
	$$\mathtt{inr} : B \rightarrow A+B$$
	One can guess the induction principle
	$$\mathtt{ind}_{A+B} : \prod_{C : A+B \rightarrow U} \Big(\prod_{a : A}C(\mathtt{inl}\;a)\Big) \rightarrow \Big(\prod_{b : B}C(\mathtt{inr}\;b)\Big) \rightarrow \prod_{x : A+ B}C(x)$$
	$$\mathtt{ind}_{A+B}(C, f_A, f_B, \mathtt{inl}\; a) :\equiv f_A\; a$$
	$$\mathtt{ind}_{A+B}(C, f_A, f_B, \mathtt{inr}\; b) :\equiv f_B\; b$$
	\item Given $x, y : A$ we have the \textbf{identity type} $x =_A y : \mathcal{U}$ (we ommit the index $A$ whenever it's easily deduced). 
	An element of this type amounts to a proof that $x$ and $y$ are equal and we call the element a path between $x$ and $y$. 
	Say $x$ and $y$ are judgmentally equal, $x \equiv y$. 
	This is captured by the element $\mathtt{idp}_x : x =_A y$. 
	For every $x  :A$ we have $\mathtt{idp}_x : x=_A x$. 
	The relevant induction principle describes how we can use elements of an identity type
	$$\mathtt{ind}_{=_{A}} : \prod_{C : \prod_{x,y : A} (x =_A y) \rightarrow \mathcal{U}} \Big( \prod_{x : A}C(x,x,\mathtt{idp}_x) \Big) \rightarrow \prod_{x,y : A} \prod_{(p : x =_A y)} C(x,y,p)$$
	The relevant computation rule gives us the judgmental (definitional) equality 
	$$\mathtt{ind}_{=_A} (C,c,x,x,\mathtt{idp_x}) \equiv c\; x$$
	We can concatenate those paths whose domains and codomains allow for it. 
	Paths form an equivalence relation. That is, if $p : x = y$ and $p' : y = z$ are such paths then we can concatenate them $p \cdot p' : x = z$. 
	We can also provide the inverse $p^{-1} : y = x$ for which we have in turn a path $q : (p\cdot p^{-1}) =_{x=x} (\mathtt{idp}_x)$ between $p\cdot p^{-1}$ and $\mathtt{idp}_x$ and another one between $p^{-1}\cdot p$ and $\mathtt{idp}_y$.

	This would be a good place to talk a bit more about the closure properties of judgmental/definitional equality $\equiv$. 
	Essentially, we demand that $\equiv$ is a congruence relation built upon those judgmental equalities we introduce elsewhere, as part of enriching our type theory with type formers, in addition to being closed under these:
	\begin{itemize}
	\item For terms $\lambda x.\;t$ and $u$ such that the application $(\lambda x.\;t)\;u$ is legal according to their typings, we have $(\lambda x.\;t)\;u \equiv t[u/x]$. 
	\item If $t \equiv t'$ and $s \equiv s'$ and the application $t(s)$ is legal, then $t(s) \equiv t'(s')$. 
	\item Closed under lambda abstraction, i.e. if $t \equiv t'$ then $\lambda x.\;t \equiv \lambda x.\;t'$. 
	\end{itemize}
	After closing under the above, we drop from $\equiv$ all pairs of terms that belong to (judgmentally) unequal types.

	We sometimes introduce judgmental equalities with a colon on the left $:\equiv$. 
	This serves merely as an indication of `direction' in the definition wherein the left hand side is being introduced as a term while it's well understood that the right hand side is a term of type theory. 
	Sometimes the lhs is just a name, a shorthand, for the rhs. 
	The colon serves no \textit{formal} purpose and the reader may ignore it, if they do not find it helpful.

@@

	Given mere proposition $B$ and $f : A \rightarrow B$, the recursion principle gives us $g : \|A\| \rightarrow B$ such that $g(|a|) \equiv f(a)$ for all $a : A$.
	This recursion principle will prove itself an indispensable tool in the sections to come. 
	Whenever, in the midst of a proof, the current goal $B$ is a mere proposition and we have access to some witness $a : \lVert A \rVert$, we are allowed by the recursion principle to assume that we have $a' : A$ and use it to construct a proof of $B$.
	\item We have an \textbf{empty} type $\bot : \mathcal{U}$ with the induction principle
	$$\mathtt{ind}_\bot : \prod_{A : \mathcal{U}}\prod_{C : A \rightarrow \mathcal{U}} \prod_{x  :\bot} C\;x$$
	\item We have a unit type $\mathbf{1} : \mathcal{U}$ for which we have a term $* : \mathbf{1}$ and a witness of $\prod_{x,y : \mathbf{1}} x =y$. 
	\item We have the $\mathbf{2}$ type with terms $0 : \mathbf{2}$ and $1 : \mathbf{2}$ and the induction principle
	$$\mathtt{ind}_\mathbf{2} : \prod_{C : \mathbf{2} \rightarrow \mathcal{U}}C(0)\rightarrow C(1) \rightarrow \prod_{x : \mathbf{2}}C(x)$$
	$$\mathtt{ind}_\mathbf{2}()$$
	\item We have the type of \textbf{natural numbers} $\mathbb{N} : \mathcal{U}$ equiped with $0 : \mathbb{N}$ and $\mathtt{succ} : \mathbb{N} \rightarrow \mathbb{N}$. 
	There is also the induction principle
	$$\mathtt{ind}_\mathbb{N} : \prod_{C : \mathbb{N} \rightarrow \mathcal{U}}C(0)\rightarrow \big(\prod_{ : \mathbb{N}} C(n) \rightarrow C(\mathtt{succ}\;n)\big)\rightarrow \prod_{n : \mathbb{N}}C\;n$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, 0) :\equiv f_0$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, \mathtt{succ}\;n) :\equiv f_s\big(n, \mathtt{ind}_\mathbb{N}(C,f_0, f_s, n)\big)$$

\end{itemize}
This concludes our informal presentation of the primitives of Type Theory. 
We will continue with definitions and results using the tools we laid out above. 
\begin{rem}
Sometimes, when constructing a term of some type, we say that we \textit{take cases} on it or some other type. 
By this we mean that we are invoking a type's induction or recursion principle. 
As an example, suppose we have $A, B, A+B : \mathcal{U}$ and $C : A +B \rightarrow \mathcal{U}$ and want to construct a term of $\prod_{x :A+B} C(x)$. 
We can do so by taking cases on $A+B$, i.e. provide an $f_A : \prod_{a :A}C(\mathtt{inl}\; a)$ which `describes where in $C$ we send elements of $A$' and an $f_B : \prod_{b :B}C(\mathtt{inr}\; b)$. 
We have effectively made use of the induction principle of the coproduct. 
\end{rem}
\begin{rem}
We sometimes say that we can decide whether $A+B$. 
This is just a shorthand for `we can construct a witness of $A+B$' and is inspired by the saying `we can decide P' where $P$ is a mere proposition and we mean that we can prove $P + (P \rightarrow \bot)$ which would also be a mere proposition. 
\end{rem}
\begin{rem}
Induction principles might appear to give us access only to dependent functions. 
They actually give us access to non-dependent functions too. 
Let $C : A \rightarrow \mathcal{U}$ be the family of types that forms the codomain of the resulting function of the induction principle. 
If for all $a : A$, $C(a) \equiv B$ where $B  :\mathcal{U}$ then we can form a non-dependent function $f : A \rightarrow B$. 
\end{rem}

\begin{lem}
Applying any function $f : A\rightarrow B$ to $a_1, a_2 : A$ such that $p : a_1 =_A a_2$ gives us a path $\mathtt{ap}_f (p) : f(a_1) =_B f(a_2)$. 
We name this \textbf{action on paths} and this instance, action of $f$ on $p$.\\
Formally,
$$\mathtt{ap} : \prod_{A,B : \mathcal{U}}\prod_{f : A \rightarrow B}\prod_{a_1,a_2 : A}(a_1 =_A a_2)\rightarrow \big(f(a_1) =_B f(a_2)\big)$$
\end{lem}
\begin{proof}
Assume the hypotheses $A,B,f, a_1, a_2$ and $p : a_1 =_A a_2$ as above. 
We need a witness of $f(a_1) =_B f(a_2)$. 
We use induction on the identity type $a_1 =_A a_2$. 
So we have to show that for $\mathtt{idp}_{a_1} : a_1 =_A a_2$, we have $f(a_1) =_B f(a_2)$. 
By $\mathtt{idp}_{a_1}$ we have $a_1 \equiv a_2$ which means we can rewrite the goal as $f(a_1) =_B f(a_1)$, for which we have a witness in the form of $\mathtt{idp}_{f(a_1)}$. 
\end{proof}
\section{Equivalence}
	Before going over what it means for two types to be equivalent, we have to work out some preliminary notions. 
	\begin{defn}
	Let $f,g : \prod_{a:A}B(a)$ where $B :A \rightarrow \mathcal{U}$. 
	We call a function of the following type a \textbf{homotopy} between $f$ and $g$
	$$f\sim g :\equiv \prod_{a:A}f(a) = g(a)$$
	\end{defn}

>>>>

\usepackage{tikz-cd}
%\usepackage{answers}
\usepackage[backend=biber,style=alphabetic,citestyle=alphabetic]{biblatex}
\addbibresource{bibliography.bib}
\usepackage{setspace}
%\usepackage{graphicx}
\usepackage{enumitem}
%\usepackage{tikz-cd}
%\usepackage{todonotes}
%\usepackage{multicol}
%\usepackage{mathrsfs}
%\usepackage[margin=1in]{geometry} 
\usepackage{hyperref}% http://ctan.org/pkg/hyperref
\usepackage{stmaryrd}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{cleveref}% http://ctan.org/pkg/cleveref
\usepackage{amssymb}
\usepackage{siunitx}
\usepackage{textcomp}
%\usepackage{ wasysym }
\usepackage{bookmark}
%\usepackage{accents}
\usepackage[parfill]{parskip}
\usepackage{appendix}

% \newcommand{\N}{\mathbb{N}}
% \newcommand{\Z}{\mathbb{Z}}
% \newcommand{\R}{\mathbb{R}}
% \newcommand{\nin}{\not \in}

% \DeclareMathOperator{\sech}{sech}
% \DeclareMathOperator{\csch}{csch}

% \newenvironment{theorem}[2][Theorem]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
% \newenvironment{definition}[2][Definition]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
% \newenvironment{proposition}[2][Proposition]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
% \newenvironment{lemma}[2][Lemma]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
% \newenvironment{exercise}[2][Exercise]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
% \newenvironment{solution}[2][Solution]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
% \newenvironment{problem}[2][Problem]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
% \newenvironment{question}[2][Question]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}
% \newenvironment{corollary}[2][Corollary]{\begin{trivlist}
% \item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2.}]}{\end{trivlist}}

% \usepackage{xcolor}
% \newcommand\todo[1]{\textcolor{red}{#1}}

@@

\newtheorem{thm}{Theorem}[chapter]
\newtheorem{lem}[thm]{Lemma}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{axiom}[thm]{Axiom}
\newtheorem{defn}[thm]{Definition}
\theoremstyle{definition}
\newtheorem{rem}[thm]{Remark}
\newtheorem{ill}[thm]{Illustration}
\newtheorem{exmp}[thm]{Example}

\crefname{prop}{proposition}{propositions}
\Crefname{prop}{Proposition}{Propositions}
\crefname{cor}{corollary}{corollaries}
\Crefname{cor}{Corollary}{Corollaries}
\crefname{lem}{lemma}{lemmas}
\Crefname{lem}{Lemma}{Lemmas}
\crefname{thm}{theorem}{theorems}
\Crefname{thm}{Theorem}{Theorems}
\crefname{axiom}{axiom}{axioms}
\Crefname{axiom}{Axiom}{Axioms}
\usepackage{illcmolthesis}
\begin{document}
 %\pagenumbering{gobble}
% --------------------------------------------------------------
%                         Start here
% --------------------------------------------------------------

\title{Lifschitz Realizability for Homotopy Type Theory}%replace with the appropriate homework number
\author{Dimitrios Koutsoulis} %if necessary, replace with your course title
\birthdate{May 27, 1992}
\birthplace{Marousi, Attica, Greece}
\defensedate{November 26, 2019}
\supervisor{Andrew Swan}
\supervisor{Benno van den Berg}
\committeemember{Benno van den Berg (co-supervisor)}
\committeemember{Nick Bezhanishvili}
\committeemember{Jaap van Oosten}
\committeemember{Andrew Swan (supervisor)}
\committeemember{Yde Venema (chair)}

%\begin{titlepage}
%\hypersetup{pageanchor=false}
{\parskip=0pt
\maketitle}
%\thispagestyle{empty}
%\end{titlepage}
%\hypersetup{pageanchor=true}
%Below is an example of the problem environment
\chapter*{Abstract}
This thesis explores a potential development of the constructive tradition of Russian Mathematics (RUSS) inside Homotopy Type Theory (HoTT). 
A short introduction to Type Theory is provided. 
Fragments of RUSS are then formalized inside it. 
A construction that follows closely Van Oosten's generalization of Lifschitz' Realizability is carried out to culminate with a consistency result; that of our selection of RUSS axioms under HoTT.

\chapter{Type Theory}
%\section{Introduction}
Martin-L\"of's intensional \textit{Type Theory} is a formal language and deductive system, that is self sufficient in the sense that it need not be formulated as a collection of axioms on top of some other formal system like First Order Logic, instead its deductive system can be built on top of its own formal language. 
The version of it that we will be considering in this text is the one used in \cite{hottbook}. 
This chapter presents, in a concise manner, definitions and results found in chapters 1-5 of \cite{hottbook}.

Central to Type Theory is the notion of \textit{Type}. Every term $a$ in Type Theory we come across, must lie in some type $A$, which we denote as $a : A$. 
Types themselves are terms of \textit{Universes}, which are types and terms themselves. 
To avoid impredicativity we assume a countable hierarchy of universes $\mathcal{U}_1,\mathcal{U}_2,\mathcal{U}_3,\ldots$ which is cummulative i.e. every universe includes all previous universes and their types. 
While working in type theory we usually simplify our view of this hierarchy and pick some $\mathcal{U}$, of arbitrary index that we do not specify, to use as our workspace. 
%Note that the relation $:$ is transitive, so $a : A$ and $A : B$ imply $a : B$. 
%This induces the cummulative hierarchy shape of the universe of types $\mathcal{U}$, wherein every level of the universe includes all lower levels and their types.

@@

\section{Type Construction Operations}\label{informalTT}
%This section constitutes an informal treatment of Type Theory. 
In this section we go over all the types of Type Theory we need. 
When working out proofs in TT, we use the constructs presented here. 
Nevertheless, this presentation is meant to reflect a formal system like the one in \Cref{formalTT}. 
We have the following list of types at our disposal.
\begin{itemize}
	\item Given types $A, B : \mathcal{U}$ we can define the type $A \rightarrow B : \mathcal{U}$ of \textbf{functions} from $A$ to $B$. 
		We can use $\lambda$-abstraction to construct elements of this type. $\lambda x. \Phi$ lies in $A \rightarrow B$ iff for $a : A$ we have $\Phi[a/x] : B$. 
		We may also right it as $\lambda (x : A). \Phi$ to make clear which type it lies in. 
		For $f : A \rightarrow B$ and $a : A$ we have that the application of $f$ on $a$, denoted as $f(a)$ or $f\;a$, lies in $B$, so $f\;a : B$.

		Functions whose type is of the form $A \rightarrow \mathcal{U}$ i.e. the codomain is $\mathcal{U}$ are called \textit{families of types}. 
		They are of special interest because they can be viewed as types themselves; types indexed by terms of the domain type. 
		So if $B : A \rightarrow \mathcal{U}$ is such a family of types, then its inhabitants would be the collection of inhabitants of all $B(a)$'s for all $a  : A$. 
		This last sentence is not sanctioned by formal type theory and only serves to make the introduction of the first dependent type, the \textit{dependent function type}, easier.
	\item Given some type $A : \mathcal{U}$ and a family of types $B$ over $A$, $B : A \rightarrow \mathcal{U}$, we have the type of \textbf{dependent functions} or \textbf{dependent products} $$\prod_{a : A} B(a)$$
	where for $f : \prod_{a : A} B(a)$ and $x : A$ we have $f(x) : B(x)$. 
	As in the case of non-dependent functions, we can use lambda abstraction to construct elements of a dependently-typed function type. 
	That way, $\lambda x.\Phi$ lies in $\prod_{a:A}B(a)$ iff $B$ is of the form $\lambda y.\Psi$ and for all $a:A$ we have $\Phi[a/x] : \Psi[a/y]$. 
	\item Given $A, B : \mathcal{U}$ we can define the \textbf{product} type $A \times B : \mathcal{U}$. 
	For $a : A$ and $b : B$ we have the pair $(a,b) : A \times B$. 
	We also have the projection functions 
	$$\mathtt{pr}_1 : A \times B \rightarrow A : (a,b) \mapsto a$$
	$$\mathtt{pr}_2 : A \times B \rightarrow B : (a,b) \mapsto b$$
	We sometimes use the notation $x.\mathtt{pr}_i$ to refer to $\mathtt{pr}_i\; x$. 
	We also make use of the alias $\mathtt{fst}$ for $\mathtt{pr}_1$ and $\mathtt{snd}$ for $\mathtt{pr}_2$.

@@

	The induction principle is the following
	$$\mathtt{ind}_{\sum_{a:A}B(a)} : \prod_{C : \sum_{a:A}B(a) \rightarrow \mathcal{U}}\Big( \prod_{a : A}\prod_{b : B(a)}C((a,b)) \Big)\prod_{x : \sum_{a:A}B(a)}C(x)$$
	$$\mathtt{ind}_{\sum_{a:A}B(a)}(C, f,(a,b) ) :\equiv f(a)(b)$$
	\item Given $A, B : \mathcal{U}$ we can construct the \textbf{coproduct} type $A + B$. 
	We can construct elements of $A + B$ using the functions
	$$\mathtt{inl} : A \rightarrow A+B$$
	$$\mathtt{inr} : B \rightarrow A+B$$
	One can guess the induction principle
	$$\mathtt{ind}_{A+B} : \prod_{C : A+B \rightarrow U} \Big(\prod_{a : A}C(\mathtt{inl}\;a)\Big) \rightarrow \Big(\prod_{b : B}C(\mathtt{inr}\;b)\Big) \rightarrow \prod_{x : A+ B}C(x)$$
	$$\mathtt{ind}_{A+B}(C, f_A, f_B, \mathtt{inl}\; a) :\equiv f_A\; a$$
	$$\mathtt{ind}_{A+B}(C, f_A, f_B, \mathtt{inr}\; b) :\equiv f_B\; b$$
	\item Given $x, y : A$ we have the \textbf{identity type} $x =_A y : \mathcal{U}$ (we omit the index $A$ whenever it's easily deduced). 
	An element of this type amounts to a proof that $x$ and $y$ are equal and we call the element a path between $x$ and $y$. 
	%Say $x$ and $y$ are judgmentally equal, $x \equiv y$. 
	%This is captured by the element $\mathtt{idp}_x : x =_A y$. 
	For every $x  :A$ we have $\mathtt{idp}_x : x=_A x$. 
	The relevant induction principle describes how we can use elements of an identity type
	$$\mathtt{ind}_{=_{A}} : \prod_{C : \prod_{x,y : A} (x =_A y) \rightarrow \mathcal{U}} \Big( \prod_{x : A}C(x,x,\mathtt{idp}_x) \Big) \rightarrow \prod_{x,y : A} \prod_{(p : x =_A y)} C(x,y,p)$$
	The relevant computation rule gives us the judgmental (definitional) equality 
	$$\mathtt{ind}_{=_A} (C,c,x,x,\mathtt{idp_x}) \equiv c\; x$$

@@

	Given mere proposition $B$ and $f : A \rightarrow B$, the recursion principle gives us $g : \|A\| \rightarrow B$ such that $g(|a|) \equiv f(a)$ for all $a : A$.
	This recursion principle will prove itself an indispensable tool in the sections to come. 
	Whenever, in the midst of a proof, the current goal $B$ is a mere proposition and we have access to some witness $a : \lVert A \rVert$, we are allowed by the recursion principle to assume that we have $a' : A$ and use it to construct a proof of $B$.
	\item We have an \textbf{empty} type $\bot : \mathcal{U}$ with the induction principle
	$$\mathtt{ind}_\bot : \prod_{A : \mathcal{U}}\prod_{C : A \rightarrow \mathcal{U}} \prod_{x  :\bot} C\;x$$
	\item We have a unit type $\mathbf{1} : \mathcal{U}$ for which we have a term $* : \mathbf{1}$ and a witness of $\prod_{x,y : \mathbf{1}} x =y$. 
	\item We have the $\mathbf{2}$ type with terms $0 : \mathbf{2}$ and $1 : \mathbf{2}$ and the induction principle
	$$\mathtt{ind}_\mathbf{2} : \prod_{C : \mathbf{2} \rightarrow \mathcal{U}}C(0)\rightarrow C(1) \rightarrow \prod_{x : \mathbf{2}}C(x)$$
	$$\mathtt{ind}_\mathbf{2}(C, c_0, c_1, 0) \equiv c_0$$
	$$\mathtt{ind}_\mathbf{2}(C, c_0, c_1, 1) \equiv c_1$$
	\item We have the type of \textbf{natural numbers} $\mathbb{N} : \mathcal{U}$ equiped with $0 : \mathbb{N}$ and $\mathtt{succ} : \mathbb{N} \rightarrow \mathbb{N}$. 
	There is also the induction principle
	$$\mathtt{ind}_\mathbb{N} : \prod_{C : \mathbb{N} \rightarrow \mathcal{U}}C(0)\rightarrow \big(\prod_{ : \mathbb{N}} C(n) \rightarrow C(\mathtt{succ}\;n)\big)\rightarrow \prod_{n : \mathbb{N}}C\;n$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, 0) :\equiv f_0$$
	$$\mathtt{ind}_\mathbb{N}(C, f_0, f_s, \mathtt{succ}\;n) :\equiv f_s\big(n, \mathtt{ind}_\mathbb{N}(C,f_0, f_s, n)\big)$$
\begin{rem}
Sometimes, when constructing a term of some type, we say that we \textit{take cases} on it or some other type. 
By this we mean that we are invoking a type's induction or recursion principle. 
As an example, suppose we have $A, B, A+B : \mathcal{U}$ and $C : A +B \rightarrow \mathcal{U}$ and want to construct a term of $\prod_{x :A+B} C(x)$. 
We can do so by taking cases on $A+B$, i.e. provide an $f_A : \prod_{a :A}C(\mathtt{inl}\; a)$ which `describes where in $C$ we send elements of $A$' and an $f_B : \prod_{b :B}C(\mathtt{inr}\; b)$. 
We have effectively made use of the induction principle of the coproduct. 
\end{rem}
\begin{rem}
We sometimes say that we can decide whether $A+B$. 
This is just a shorthand for `we can construct a witness of $A+B$' and is inspired by the saying `we can decide P' where $P$ is a mere proposition and we mean that we can prove $P + (P \rightarrow \bot)$ which would also be a mere proposition. 
\end{rem}
\begin{rem}
Induction principles might appear to give us access only to dependent functions. 
They actually give us access to non-dependent functions too. 
Let $C : A \rightarrow \mathcal{U}$ be the family of types that forms the codomain of the resulting function of the induction principle. 
If for all $a : A$, $C(a) \equiv B$ where $B  :\mathcal{U}$ then we can form a non-dependent function $f : A \rightarrow B$. 
\end{rem}
\section{Judgmental Equality}
This would be a good place to talk a bit more about the closure properties of judgmental/definitional equality $\equiv$. 
	Essentially, we demand that $\equiv$ is a congruence relation built upon those judgmental equalities we introduce elsewhere, as part of enriching our type theory with type formers, in addition to being closed under these:
	\begin{itemize}
	\item For any term $a : A$ we have $a \equiv a$. 
	\item For terms $\lambda x.\;t$ and $u$ such that the application $(\lambda x.\;t)\;u$ is legal according to their typings, we have $(\lambda x.\;t)\;u \equiv t[u/x]$. 
	\item If $t \equiv t'$ and $s \equiv s'$ and the application $t(s)$ is legal, then $t(s) \equiv t'(s')$. 
	\item Closed under lambda abstraction, i.e. if $t \equiv t'$ then $\lambda x.\;t \equiv \lambda x.\;t'$. 
	After closing under the above, we drop from $\equiv$ all pairs of terms that belong to (judgmentally) unequal types. 
	\begin{rem}
	We sometimes introduce judgmental equalities with a colon on the left $:\equiv$. 
	This is just a stylistic choice, i.e. the colon can be ignored. 
	\end{rem}
	% This serves merely as an indication of `direction' in the definition wherein the left hand side is being introduced as a term while it's well understood that the right hand side is a term of type theory. 
	% Sometimes the lhs is just a name, a shorthand, for the rhs. 
	% The colon serves no \textit{formal} purpose and the reader may ignore it, if they do not find it helpful. 
	\begin{rem}
	This liberty is justified, as one would expect, by the inclusion of relevant inference rules in the underlying formal system of Type Theory. 
	\end{rem}
	% With the above facts about judgmental equality in mind, along with the induction principle of equality, we can do the following in TT. 
	% We can concatenate those paths whose domains and codomains allow for it. 
	% That is, if $p : x = y$ and $p' : y = z$ are such paths then we can concatenate them $p \cdot p' : x = z$, the proof of this consists of invoking $\mathtt{ind}_{=}$ twice, once for $p$ and once for $p'$, then we have to prove $x=z$ where $x\equiv y$ and $y \equiv z$, a trivial result. 
	% We can also provide the inverse $p^{-1} : y = x$ for which we have in turn a path $q : (p\cdot p^{-1}) =_{x=x} (\mathtt{idp}_x)$ between $p\cdot p^{-1}$ and $\mathtt{idp}_x$ and another one between $p^{-1}\cdot p$ and $\mathtt{idp}_y$. 
	% We conclude that paths form an equivalence relation. 
	This concludes our informal presentation of the primitives of Type Theory. 
	We will continue with definitions and results using the tools we laid out above. 
	\section{On paths}
	In this section we go over some results about manipulating identity paths. 
	\begin{lem}
	We can construct the following function with infix notation, which we call \textbf{concatenation of paths}
	$$\_\cdot \_ : \prod_{A : \mathcal{U}}\prod_{a_1, a_2, a_3 : A}(a_1 =_A a_2) \rightarrow (a_2 =_A a_3) \rightarrow (a_1 =_A a_3)$$
